National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-05-08 analysis/paper/paper.Rmd
Make paper and scripts called by it work with updated package

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

<!-- The actual document text starts here: -->
# Abstract
Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian integrated population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.
# Introduction

@@

cd <- here::here()
app_path <- gsub("Caribou-Demographic-Projection-Paper","BayesianCaribouDemographicProjection",cd,fixed=T)

@@

runScenarios = F 
doPriorCalibration=F #Note the prior calibration is time consuming. 
```
TO DO: build large table of all parameters as the various sections are written. Adjust UI for consistent terminology along the way.

@@

## A simple Bayesian IPM to integrate local demographic data and national disturbance-demographic relationships

Following [@eacker_web-based_2019], the observed number of calves $c_t$ is a binomially distributed function of the observed number of adult female caribou $a_t$ and the estimated recruitment probability $R_t$: $$c_t \sim R_t^{c_t}(1-R_t)^{a_t-c_t}.$$
We model recruitment probability as a function of anthropogenic disturbance and fire [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022], with a log link and Gaussian distributed random year effect [@eacker_web-based_2019]: $$\log(R_t)=\beta^R_0+\beta^R_a anthro_t+\beta^R_f fireExclAnthro_t+\epsilon^R_t; \epsilon^R_t \sim \text{normal}(0,\sigma^2_{R}).$$ 
Estimated recruitment probability is adjusted for sex ratio and delayed age of first reproduction [@eacker_web-based_2019;@decesare_estimating_2012] to get expected recruits per female: $$\tilde{R}_t=\frac{R_t/2}{1+R_t/2}.$$

Also following [@eacker_web-based_2019], Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio collar data using the survival R package [@therneau_modeling_2000], with precisions for years with no observed mortality estimated using a separate Bayesian model. Observed estimated survival is a Gaussian distributed function of the true survival probability: $$\hat{S_t} \sim \text{normal}(S_t,1/\tau_t).$$ 
We model true survival probability as an adjusted function of anthropogenic disturbance  [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022], with a log link and Gaussian distributed random year effect [@eacker_web-based_2019]: $$S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a anthro_t+\epsilon^S_t; \epsilon^S_t \sim \text{normal}(0,\sigma^2_{S}).$$ 
To assess sensitivity of results to the survival estimation method, we also implemented an alternative parametric exponential survival model, wherein the state of an individual animal in year $t$ and month $m$ is Bernoulli distributed (ref IPM book): $$A_{t,m} \sim \text{Bernoulli}(S^{1/12}_t A_{t,m-1}).$$

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prediction interval from the IPM matches the range between the 2.5% and 97.5% quantiles of 1000 simulated recruitment (Supplement X).

@@

knitr::kable(tbl,caption="Model priors parameters are Gaussian distributed with expected values ")
```

TO DO:

* ensure that INF_saf_INF_r_JAGS_extentTime.txt and INF_saf_INF_r_JAGS_expSurvival2.txt are identical except for the survival observation model. Or better yet, write code to sub in correct survival model more elegantly.
* note various bits of truncation in the JAGS code to ensure R and S remain between 0 and 1 (not guaranteed given log link). Not all documented in this explanation. Bayesian wizard might know a better way. Or at least know how to explain.
* note Eacker survival method gives biased results in years when <12 months of survival data is available.
* Note I am dropping survival information for animals that are not alive at the beginning of a year to deal with censoring issues. There is probably some more elegant way to deal with this, but it won't make any difference as long as the collar deployment month is set to 1.
* add references to documentation of popGrowthJohnson and other caribouMetrics functions.
* consider & comment on alternative choices made in this example https://poissonconsulting.shinyapps.io/bboushiny/. Follow Eacker example in all particulars, change some aspects, or do comparisons?

## A method for simulating example data

@@

# Appendix

>>>>

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian integrated population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.

keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

<!-- The actual document text starts here: -->
# Introduction

NOTE: find working version of intro and discussion [here:] (https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true).

@@

@@

runScenarios = F 
doPriorCalibration=F #Note the prior calibration is time consuming. 
```

@@

## Integration of local demographic data and national disturbance-demographic relationships in Bayesian population model

Following Eacker et al. [-@eacker_web-based_2019], the observed number of calves $\hat{C}_t$ is a binomially distributed function of the observed number of adult female caribou $\hat{W}_t$ and the estimated recruitment rate $R_t$: $$\hat{C}_t \sim \text{Binomial}(\hat{W}_t,R_t).$$

We model recruitment probability as a function of anthropogenic disturbance and fire with a log link [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$\log(R_t)=\beta^R_0+\beta^R_a a_t+\beta^R_f f_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).$$ 
Estimated recruitment probability is adjusted for sex ratio to get expected recruits per female: $\tilde{R}_t=R_t/2$. We include an option to also adjust for delayed age of first reproduction [@eacker_web-based_2019;@decesare_estimating_2012], but to maintain consistency with Johnson et al. [-@johnson_science_2020] we do not apply this adjustment by default: $$\tilde{R}_t=\frac{R_t/2}{1+R_t/2}.$$

Also following [@eacker_web-based_2019], Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio collar data using the survival R package [@therneau_modeling_2000], with precisions for years with no observed mortality estimated using a separate Bayesian model. Observed estimated survival is a Gaussian distributed function of the true survival probability: $$\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).$$ 
We model true survival probability as an adjusted function of buffered anthropogenic disturbance $a_t$ with a log link [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a a_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).$$ 
To assess sensitivity of results to the survival estimation method, we also implemented an alternative parametric exponential survival model, wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed (ref IPM book): $$\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).$$

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prediction interval from the IPM matches the range between the 2.5% and 97.5% quantiles of 1000 simulated recruitment (Appendix A).

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters.")
#TO DO: indicate which of these are from Johnson, and which are calibrated.
```

We assume a simple two-class demographic model, with a census in late winter of surviving females and new recruits to the post-juvenile (aged 1+) class. The number of post-juvenile females that survive from year $t$ to the census $A_t$ is binomially distributed with survival probability $S_t$: $A_{t} \sim \text{Binomial}(N_t,S_t)$. The number of juveniles recruiting to the post-juvenile class at the census is a Poisson distributed function of the number of surviving post-juvenile females and the adjusted recruitment rate $\tilde{R}_t$: $J_{t} \sim \text{Poisson}(\tilde{R}_tA_t)$. The post-juvenile female population in the next year includes both survivors and new recruits: $N_{t+1}=A_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0 when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more stable estimates of population status, we report average population growth over a period of $y$ years, calculated as $\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial post-juvenile population size $N_0$ is 1000 animals. For reference, all variables and parameters are defined in Table A1.

<!--
Annotated equations for poster.
$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{A_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
$$A_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al. 2020}}=\beta^S_0+ \underbrace{\beta^S_aa_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
$$J_{t} \sim \text{Poisson}(A_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a a_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_ff_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->

TO DO: 
* note distribution of random effect parameters are truncated to ensure R and S remain within reasonable bounds (not guaranteed given log link). Not yet documented here.
* note Eacker survival method gives biased results in years when <12 months of survival data is available.
* Note I am dropping survival information for animals that are not alive at the beginning of a year to deal with censoring issues. There is probably some more elegant way to deal with this, but it won't make any difference as long as the collar deployment month is set to 1.
* add references to documentation of popGrowthJohnson and other caribouMetrics functions.
* consider & comment on alternative choices made in this example https://poissonconsulting.shinyapps.io/bboushiny/. Follow Eacker example in all particulars, change some aspects, or do comparisons?
* Ensure notation consistency amongst manuscript, UI, documentaiton and code where appropriate.

## A method for simulating example local demographic data

@@

# Appendix
##Definitions of variables and parameters in the Bayesian integrated population model

Table A1. Variables and parameters in the Bayesian integrated population model.

|**Symbol**|**Units**|**Description**|
|:-----|-----|:-----------|
|**Indices & constants**|
|$t$|year|Year.|
|$m$|month|Month.|
|$i$|ID|Individual animal identifier.|
|$y$|years|Assessment period for population growth rate.|
|$N_0$|animals|Initial adult population size.|
|**Observed**|
|$a_t$|%|Buffered anthropogenic disturbance.|
|$f_t$|%|Fire disturbance.| 
|$\hat{C}_t$|animals|Calves observed in calf:cow survey.|
|$\hat{W}_t$|animals|Cows observed in calf:cow survey.|
|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
|**Reported state**|
|$R_t$|rate|Recruitment (calves per cow).|
|$S_t$|probability|Adult female survival.|
|$N_t$|animals|Adult female population size.|
|$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
|**Other state**|
|$\tilde{R}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$A_t$|animals|Surviving adult females before recruitment.|
|$J_t$|animals|New recruits to adult age class.|
|$\lambda_t$|rate|Annual population growth.|
|**Parameters**|
|$\beta^R_0$||Recruitment intercept.|
|$\beta^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\beta^R_f$||Fire slope for recruitment.|
|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|
|$\beta^S_0$||Survival intercept.|
|$\beta^S_a$||Anthropogenic disturbance slope for survival.|
|$\sigma^2_{S}$||Variance in random effect of year on survival.|
